Validation of Visual Statistical Inference, Applied to Linear Models

### 2012-07-18 paper_jasa.tex
done proof-reading

\subsection{Model fitting}
\noindent For each parameter combination, the effect is computed as

@@

\noindent Figure \ref{fig:power_loess_effect} shows an overview of estimated power  against effect for the three experiments.  Responses from each experiment are summarized by effect size and represented as dots,  with size indicating the number of responses.
A loess  fit to the data gives an estimate of  the observed power for different effect sizes, with grey bands indicating simultaneous bootstrap confidence bands \citep{buja}. For comparison, the dashed lines show the corresponding power curves of the conventional tests. Not surprisingly, the observed power for experiments 1 and 2 is lower than the power of the conventional test.  It is encouraging to see that visual inference mirrors the power vs effect relationship of conventional testing, and comes quite close in experiment 2, where scatterplots are used as the test statistic. 
Note that for effect $E=0$, the power is close to 0.05 (Type-I error) for both experiments 1 and 2, making the fixed offset a reasonable assumption.

@@

%Different subjects are expected to have different skill levels in evaluating lineups. This is accounted for in the mixed effects model (Equation \ref{eqn:mixed}).   
\noindent 
Figure  \ref{fig:power_mixed_subject} shows power curves for both the overall experiment and subject-specific variations. The thick continuous line shows overall estimated power, the thinner lines correspond to subject-specific power curves. For comparison, the  dashed lines show power curves of the conventional test. Subject-specific power is quite different between the three experiments. In experiment 1 subjects performed similarly, and substantially worse than the conventional test. In experiment 2 there is more variability between subjects, with some doing better than the conventional test on small effects. In experiment 3 there is the most subject-specific variation. Some subjects performed substantially better than the conventional test, and on average the visual test was better.

@@

\noindent One assumption made in order to calculate the effect of lineup size in the calculations of visual $p$-value and signal strength was that subjects would tend to pick the plot in the lineup that had the strongest signal. In experiments 1 and 2, this corresponds to the plot with the smallest $p$-value. This can be examined for the data collected, to see if this assumption is, indeed, reasonable. %In Section \ref{sec:size} in order to make comparisons with conventional tests an assumption was made that the subject would pick the plot in the lineup that had the lowest $p$-value. In this section we take a look at the picks that subjects made, to see if this is, indeed, true.

@@

There are some noticeable exceptions to this rule. In experiment 1, when $\beta=0, n=100, \sigma=5, rep=1$ people overwhelmingly chose a plot with much larger $p$-value, similarly, for 
parameters $\beta=5, n=300, \sigma=12, rep=3$, people tended to pick the plot with the second smallest $p$-value.
For several of these exceptions, along with several easy lineups, we followed up with a new experiment using an eye-tracker to examine what the observers were looking at in making their choices \citep{zhao:2012}.
%(Lineups chosen for the eye-tracker experiment were in experiment 1, (1) $n=100, \beta=3, \sigma=5, rep=1$ (2) $n=100, \beta=5, \sigma=5, rep=1$ (3) $n=100, \beta=10, \sigma=12, rep=1$ and (4) $n=300, \beta=5, \sigma=5, rep=1$, and in experiment 2, (5) $n=100, \beta=3.5, \sigma=12, rep=1$ (6) $n=100, \beta=4.5, \sigma=12, rep=2$ (7) $n=300, \beta=1, \sigma=5, rep=1$ and (8) $n=300, \beta=1, \sigma=5, rep=2$.)

Experiment 3 does not have $p$-values associated with each plot, so the assumption can only be tested with the first two experiments. Practically, this is not an assumption that is explicitly made, because there will not be a $p$-value available. Rather it is assumed that the subject has the ability to select the plot with the most extreme structure. A few of the lineups indicated that occasionally people cue on different patterns, for example, in experiment 1 the lineup $\beta=0, n=100, \sigma=5, rep=1$ (first row, fourth column) many subjects chose the same plot, that had a higher $p$-value. Some of these anomalous lineups were selected for follow-up studies using an eye-tracker to investigate what people were looking at.

@@

Visual inference is like other types of computationally intensive test procedures, such as permutation tests, where the statistics from the data are compared with those from null data. But unlike these approaches visual inference is constrained to make the comparison with just a few draws $(m-1)$ from the null distribution. How this small set of null plots influences the subjects choice is important for understanding the reliability of visual inference. If the actual data plot is very different from the null plots, then the null plots should not have much influence on the choice. Measuring the difference, generally, between the plots is almost impossible! But in the controlled setting here we can use $p$-value of the test statistic calculated on the data used in each plot as a proxy for similarity of structure between the plots.

We have seen that the subjects tend to pick the plot in the lineup that has the lowest $p$-value (Figures \ref{fig:P-val_log}, \ref{fig:P-val_log2}). What we are interested in here is how this pick is affected by the distributin=on of $p$-values of other plots in the lineup, particularly the $p$-value of the null plot with the strongest structure. If there is a null plot with a small $p$-value, or one close to that of the actual data plot, we would expect that subjects have a harder time detecting the actual data plot. Figure \ref{fig:pval_difference} investigates this. The difference between the $p$-value of the actual data is compared with the lowest from the null plots. This is plotted horizontally, and the proportion correct is plotted vertically. Negative values indicate lineups where the actual data plot had a smaller $p$-value than the minimum of the null plots. In experiment 1 (boxplots) there were a lot more lineups where the actual data plot had the smaller $p$-value, but only barely. This caused quite some confusion for subjects, as seen because the variability in the proportion correct is huge for these lineups. Similarly large variability in correctness can be seen in the results of experiment 2 (scatterplots) except that the greater range of of differences in $p$-values shows the strength of subject's ability to pick the plot with most structure. Figures \ref{fig:P-val_log}, \ref{fig:P-val_log2} show a little more complicated story. When there is a big difference between the $p$-values (eg experiment 1, $\beta>7$) the subjects' as a force choose the same plot. when there is less difference the distribution of counts is much more evenly spread between plots (eg experiment 1, $\beta=1$).

@@

In practice, $p$-value is not going to be a valid way to compare plots. Rather metrics that can measure how graphical elements from one plot to another are perceived similarly are needed. This is investigated in \citet{niladri:2012}. Here numerical measures of the similarity between plots are proposed to provide quality metrics for lineups.

@@

A little known error amongst statisticians is what was coined as Type III error in \citet{mosteller:48}. This occurs when the null hypothesis is rejected for thw rong reason. Experiment 3 is prone to this type of error. Observers were asked to identify the plot that had the largest slope, either positive or negative. But the actual data plot had a cluster of points, the contamination that forced the conventional test to fail to see any trend. For the human eye this cluster of points is probably as visible as the association between the remaining points, leading the observer to detect the actual data plot by looking for the cluster instead of the slope. This would be considered to be a Type III error because it is not what was requested, but it still gives the correct answer. This is not a problem with the lineup protocol, because we want observers to react to anything that sets the the actual data plot apart. Here, in the controlled but simple setting, it is of interest to see what observers cued to in the actual data plot.

@@

The question that is posed to observers can affect the abililty to evaluate lineups. In these experiments subjects were asked very specific questions, because the purpose was to compare with very specific conventional testing. In practice, the question posed will be very general to allow observers to choose plots with any remarkable features, which enables the data mining discovery process.

@@

Even though a single observer can evaluate a lineup, as suggested in the initial lineup protocol, this work recommends having multiple observers. This ensures the ability to obtain a good estimate of the $p$-value for the test, and more confidence in the decision.

Experimental subjects in the simulation study came from various level of education, both genders and many different ages. Many subjects had little background in statistics, and were not necessarily aware of statistical plots. Yet the results were good. We expect that visual inference could be a useful tool for high school students to learn about randomness as a primer to statistical thinking. A major advantage of visual statistical inference is that it is intuitive, as illustrated by the performance of subjects with limited education, and it is may make statistics fun to learn.

@@

When there is no conventional test available, visual testing procedures can be used due to its few assumptions and non-parametric nature. We have successfully used the procedures informally for local projects. It is hoped that this proves to be extremely valuable in data mining applications, and exploratory analyses, where there are no exising gauges of statistical significance. The lineup protocol operates similarly to general tests, such as two-sided, analysis of variance $F$-tests, and Wilks $\Lambda$ for multivariate analysis of variance. If the null hypothesis is rejected, it is generally that we can say that ``there is something there'' but not specifically what it is in the data that triggers the rejection. Follow-up questions on the reasons provide qualitative insight. Multiple comparisons are often done to refine and understand the results of the general conventional testing, and perhaps some similar approaches might be developed for visual inference.

>>>>

\subsection{Model fitting}
 For each parameter combination, the effect is computed as

@@

 Figure \ref{fig:power_loess_effect} shows an overview of estimated power  against effect for the three experiments.  Responses from each experiment are summarized by effect size and represented as dots,  with size indicating the number of responses.
A loess  fit to the data gives an estimate of  the observed power for different effect sizes, with grey bands indicating simultaneous bootstrap confidence bands \citep{buja}. For comparison, the dashed lines show the corresponding power curves of the conventional tests. Not surprisingly, the observed power for experiments 1 and 2 is lower than the power of the conventional test.  It is encouraging to see that visual inference mirrors the power vs effect relationship of conventional testing, and comes quite close in experiment 2, where scatterplots are used as the test statistic. 
Note that for effect $E=0$, the power is close to 0.05 (Type-I error) for both experiments 1 and 2, making the fixed offset a reasonable assumption.

@@

%Different subjects are expected to have different skill levels in evaluating lineups. This is accounted for in the mixed effects model (Equation \ref{eqn:mixed}).   
Figure  \ref{fig:power_mixed_subject} shows power curves for both the overall experiment and subject-specific variations. The thick continuous line shows overall estimated power, the thinner lines correspond to subject-specific power curves. For comparison, the  dashed lines show power curves of the conventional test. Subject-specific power is quite different between the three experiments. In experiment 1 subjects performed similarly, and substantially worse than the conventional test. In experiment 2 there is more variability between subjects, with some doing better than the conventional test on small effects. In experiment 3 there is the most subject-specific variation. Some subjects performed substantially better than the conventional test, and on average the visual test was better.

@@

 One assumption made in order to calculate the effect of lineup size in the calculations of visual $p$-value and signal strength was that subjects would tend to pick the plot in the lineup that had the strongest signal. In experiments 1 and 2, this corresponds to the plot with the smallest $p$-value. This can be examined for the data collected, to see if this assumption is, indeed, reasonable. %In Section \ref{sec:size} in order to make comparisons with conventional tests an assumption was made that the subject would pick the plot in the lineup that had the lowest $p$-value. In this section we take a look at the picks that subjects made, to see if this is, indeed, true.

@@

There are some noticeable exceptions to this rule. In experiment 1, when $\beta=0, n=100, \sigma=5, rep=1$ people overwhelmingly chose a plot with much larger $p$-value, similarly, for 
parameters $\beta=5, n=300, \sigma=12, rep=3$, people tended to pick the plot with the second smallest $p$-value. 
For several of these exceptions, along with several easy lineups, we followed up with a new experiment using an eye-tracker to examine which patterns or features participants are cueing on in making their choices \citep{zhao:2012}.
%(Lineups chosen for the eye-tracker experiment were in experiment 1, (1) $n=100, \beta=3, \sigma=5, rep=1$ (2) $n=100, \beta=5, \sigma=5, rep=1$ (3) $n=100, \beta=10, \sigma=12, rep=1$ and (4) $n=300, \beta=5, \sigma=5, rep=1$, and in experiment 2, (5) $n=100, \beta=3.5, \sigma=12, rep=1$ (6) $n=100, \beta=4.5, \sigma=12, rep=2$ (7) $n=300, \beta=1, \sigma=5, rep=1$ and (8) $n=300, \beta=1, \sigma=5, rep=2$.)

%Experiment 3 does not have $p$-values associated with each plot, so the assumption can only be tested with the first two experiments. Practically, this is not an assumption that is explicitly made, because there will not be a $p$-value available. Rather it is assumed that the subject has the ability to select the plot with the most extreme structure. A few of the lineups indicated that occasionally people cue on different patterns, for example, in experiment 1 the lineup $\beta=0, n=100, \sigma=5, rep=1$ (first row, fourth column) many subjects chose the same plot, that had a higher $p$-value. Some of these anomalous lineups were selected for follow-up studies using an eye-tracker to investigate what people were looking at.

@@

 Visual inference falls into the same framework as randomization tests, where the statistics from the data are compared with those from null data. 
 But unlike these approaches visual inference is constrained to make the comparison with just a few draws $(m-1)$ from the null distribution. How this small set of null plots influences the subjects choice is important for understanding the reliability of visual inference. If the actual data plot is very different from the null plots, then the null plots should not have much influence on the choice. Measuring the difference, generally, between  plots is almost impossible. However,  in this controlled setting we can use $p$-values of the test statistic calculated on the data used in each plot as a proxy for similarity of structure between the plots.

We have seen that the subjects tend to pick the plot in the lineup that has the lowest $p$-value (Figures \ref{fig:P-val_log}, \ref{fig:P-val_log2}). What we are also interested in  is how this pick is affected by the distribution  of $p$-values of other plots in the lineup, particularly the $p$-value of the null plot with the strongest structure. If there is a null plot with a small $p$-value, or one close to that of the actual data plot, we would expect that subjects have a harder time detecting the actual data plot. Figure \ref{fig:pval_difference} investigates this. The difference between the $p$-value of the actual data is compared with the lowest from the null plots. This is plotted horizontally, and the proportion correct is plotted vertically. Negative values indicate lineups where the actual data plot had a smaller $p$-value than the minimum of the null plots. In experiment 1 (boxplots) there were a lot of lineups where the actual data plot had the smallest $p$-value, but only just. This caused quite some confusion for subjects, as seen because the variability in the proportion correct is huge for these lineups. Similarly large variability in correctness can be seen in the results of experiment 2 (scatterplots) except that the greater range of of differences in $p$-values shows the strength of subject's ability to pick the plot with most structure. Figures \ref{fig:P-val_log}, \ref{fig:P-val_log2} shed some more light on this story: when there is a big difference between the $p$-values (eg experiment 1, $\beta>7$) the subjects' as one force choose the same plot. When there is less difference the distribution of counts is much more evenly spread between plots (eg experiment 1, $\beta=1$).

@@

In practice, the $p$-value is not going to be a valid way to compare plots. Rather metrics that can measure how graphical elements from one plot to another are perceived similarly are needed. This is investigated in \citet{niladri:2012}. Here, numerical measures of the similarity between plots are proposed to provide quality metrics for lineups.

@@

A little known error amongst statisticians is what was coined as Type III error in \citet{mosteller:48}. This occurs when the null hypothesis is correctly rejected but for the wrong reason. Experiment 3 is prone to this type of error. Observers were asked to identify the plot that had the largest slope, either positive or negative. But the actual data plot had a cluster of points, the contamination that forced the conventional test to fail to see any trend. For the human eye this cluster of points is  as visible as the association between the remaining points, leading the observer to detect the actual data plot by looking for the cluster instead of the slope. This would be considered to be a Type III error because it is not what was requested, but  still gives the correct answer. This is not a problem with the lineup protocol, because we want observers to react to anything that sets the the actual data plot apart. Here, in the controlled but simple setting, it is of interest to see what observers cued to in the actual data plot.

@@

The question that is posed to observers can affect the abililty to evaluate lineups. In these experiments subjects were asked very specific questions, because the purpose was to compare with very specific conventional testing. In practice, the question posed will be more general to allow observers to choose plots with any remarkable features, which enables the data mining discovery process.

@@

Even though a single observer can evaluate a lineup, as suggested in the initial lineup protocol, this work recommends having multiple observers to ensure a more reliable estimate of the $p$-value for the test, and more confidence in the decision.

Experimental subjects in the simulation study came from various level of education, both genders and many different ages. Many subjects had little background in statistics, and were not necessarily aware of statistical plots. Yet, the results were good. We expect that visual inference could be a useful tool for high school students to learn about randomness as a primer to statistical thinking. A major advantage of visual statistical inference is that it is intuitive, as illustrated by the performance of subjects with limited education, and it  may make statistics fun to learn.

@@

When there is no conventional test available, visual testing procedures can be used due to its few assumptions and non-parametric nature. We have successfully used the procedures informally for local projects. It is hoped that this proves to be extremely valuable in data mining applications, and exploratory analyses, where there are no existing gauges of statistical significance. The lineup protocol operates similarly to general tests, such as two-sided, analysis of variance $F$-tests, and Wilks $\Lambda$ for multivariate analysis of variance. If the null hypothesis is rejected, it is generally that we can say that ``there is something there'' but not specifically what it is in the data that triggers the rejection. Follow-up questions on the reasons provide qualitative insight. Multiple comparisons are often done to refine and understand the results of the general conventional testing, and perhaps some similar approaches might be developed for visual inference.
